---
title: "Differential analysis with DESeq2"
output: html_notebook
---

- In this analysis, /*all gene statics*/ are saved to .csv file for facilitating further analysis. 
- Also, the normalized counts were also saved. 
- The adjusted p value (FDR) threshold is set to 0.05 for identifying significant.
- Removing all the post ampicillin replicas for this analysis.

# Load package
```{r}
library("DESeq2")
library("EnhancedVolcano")
```
# Load data
```{r}
control_gene_index = as.integer(scan("HelpFiles/non_regulated_gene_index_for_R.txt", what = "integer", sep = ","))

control_gene_index[1:10]
```

```{r}
# load the data 
# round to integer
countfile = './countdata.txt'
countsdata = read.csv(countfile, sep="\t", header=TRUE,row.names=1 )
countsdata = round(countsdata, digits = 0)
head(countsdata,4)
```

# Remove OD.L50.01
```{r}
# remove the outlier sample
countsdata$OD.L50.01 = NULL
countsdata$T0.L25.10 = NULL
countsdata$T1.L01.12 = NULL
```

```{r}
dim(countsdata)
```
```{r}
countsdata[1,]
```

```{r}
for (i in range(4488)){
  if (sum(countsdata[i,]>=1)==0){print(countsdata[i,])}
}
```


# Function: Systematically rename all the columns
There were repeated names after I change the "OD" to "T0". This is really bad. 
I need to change the column names that there are not repetitive. 
I am not sure if I am going to meet similar problems. But if I do, I supposed I have some certain *pattern*, and then find this *pattern* in all my column names, and then change the column names system matically. 
```{r}
sys_naming = function(dataset, condition) {
  
  # This function take in a dataset, which should be a R dataframe, where each row is a variable, and each column is a sample name. The function will take in both the data frame and the condition names, as an indicator to return an systmatically re-named dataframe. 
  # input:
  # dataset - the count dataframe
  # condition - an named vector for conditions, where the vector are the new patterns, and the name of the vector are the patterns happend in the column names
  # output:
  # new_dataset - dataframe with systematically re-named columns
  
  name_vector = colnames(dataset)    # obtain the column names
  
  new_name = substr(name_vector,1,6)    # for loop for replacing all the condition names
  for ( i in 1:length(condition)) {    
    new_name = gsub(names(condition[i]), condition[[i]], new_name)
  }
    
  num_string = gsub(" ", "0", format(1:length(name_vector), trim = FALSE))    # column numbering
  new_name_vector = paste(new_name, num_string, sep = "")    # the new column names
  
  names(name_vector) = new_name_vector    # named vector that index the new name will trace back to its old names
  #print(name_vector)
  
  return(name_vector)
}
```

```{r}
# use the function sys_naming for changing the dataframe column names. 
conditions = c("pre_lac01_", "pre_lac25_", "pre_lac50_", "pst_lac01_", "pst_lac50_", "pre_lac01_", "pre_lac25_", "pre_lac50_")
names(conditions) = c("T0.L01", "T0.L25", "T0.L50", "T1.L01", "T1.L50", 
                      "OD.L01", "OD.L25", "OD.L50")
new_con_names = sys_naming(countsdata, conditions)

counts_data_qc = countsdata
colnames(counts_data_qc) = names(new_con_names)    # change dataframe column names
colnames(counts_data_qc)
```

# Create handle dataframe
```{r}
sample = colnames(counts_data_qc)
handle = data.frame(row.names = sample)

handle$condition_all = substr(sample, 1,9)
handle$lactose = substr(handle$condition_all,5,9)

#handle$lactose = gsub("lac01", "0.1 mg/ml - low",handle$lactose)
#handle$lactose = gsub("lac25", "2.5 mg/ml - mid",handle$lactose)
#handle$lactose = gsub("lac50", "50 mg/ml - high",handle$lactose)

handle$ampicillin = substr(handle$condition_all,1,3)

View(handle)
all(rownames(handle)==colnames(countsdata))
```


```{r}
dim(countsdata)
```

# Differential analysis

Then let us define the differential analysis object.
Note here, the design formula. I suppose it is a GLM for addation of all conditions. It's a good idea to consider other design.

Perform the differential analysis 

```{r}
reflist = c("pre_lac01","pre_lac25", "pre_lac50")
```

### Function: vocanoplot

```{r}
# volcano plot
# desfolder - folder for containing the files

myvocano = function(result_matrix, result_condition, desfolder, refhere, p_bar = 0.05){
  
  # reorder the results according to the p value
  resOrdered <- result_matrix[order(result_matrix$pvalue),]
  
  # create a subpath for reference condition in the desfolder
  subdir = paste(desfolder,"/",refhere, sep = "")
  if (dir.exists(subdir) == FALSE) {dir.create(subdir)}
  
  # define pdf path
  plotname = paste(subdir, "/", substr(result_condition,15,36),".pdf", sep = "")
  pdf(plotname, width=5.5, height=6)
  # select significant DE data
  result_sig <- subset(resOrdered, padj < p_bar)
  vocano = EnhancedVolcano(
    result_sig, 
    lab = rownames(result_sig), 
    #selectLab = result_sig@rownames[1:30],
    selectLab = c(),
    pCutoff = p_bar,
    FCcutoff = 2,
    title = substr(result_condition,15,30),
    x = "log2FoldChange", y = "pvalue",
    xlab = bquote(~Log[2]~ "fold change"), ylab = bquote(~-Log[10]~italic(FDR))
    )
  print(vocano)
  dev.off()
}
```

```{r}
# print and save summary log
summarylog = function(result_matrix, result_condition, desfolder, refhere, p_bar = 0.05, getall=0){

  summary(result_matrix, alpha = p_bar)
  
  # create a subpath for reference condition in the desfolder
  subdir = paste(desfolder,"/",refhere, sep = "")
  if (dir.exists(subdir) == FALSE) {dir.create(subdir)}

  # export summary to a txt file
  result_summary <- capture.output(summary(result_matrix, alpha = p_bar))
  cat(
    result_summary, 
    file=paste(subdir, "/", substr(result_condition,15,36), "_summary", ".txt", sep = ""),
    sep="\n", append=FALSE)
}

# export results as .csv file
resultcsv = function(result_matrix, result_condition, desfolder, refhere, p_bar = 0.05, getall=0){
  
  # getall = 0, then .csv record only significant genes
  # getall = 1, then .csv record the entire genes records
  
  # We can order our results table by the smallest p value
  resOrdered <- result_matrix[order(result_matrix$pvalue),]
  
  # export data
  if (getall==1){result_sig <- subset(resOrdered)}
  else {result_sig <- subset(resOrdered, padj < p_bar)}
  # create a subpath for reference condition in the desfolder
  subdir = paste(desfolder,"/",refhere, sep = "")
  if (dir.exists(subdir) == FALSE) {dir.create(subdir)}
  
  # define pdf path
  csvname = paste(subdir, "/", substr(result_condition,15,36), ".csv", sep = "")
  print(csvname)
  write.csv(as.data.frame(result_sig), file=csvname)
}
```


```{r}
deseqloop = function(dataobject, reflist, vocadir, csvdir){
  dds_de_list = list()
  for (i in c(1:length(reflist))) {
    # factor function is available for aimed comparisions
    print(reflist[i])
    dataobject$condition_all <- relevel(dataobject$condition_all, ref = reflist[i])
    
    # DE analysis
    #dds_de = DESeq(dataobject, test = "Wald")
    dds_de <- estimateSizeFactors(dataobject, controlGenes = control_gene_index)
    #dds_de <- estimateSizeFactors(dataobject)
    dds_de <- estimateDispersions(dds_de)
    dds_de <- nbinomWaldTest(dds_de)
    
    dds_de_list[i]=dds_de
    
    # print results names
    renames = resultsNames(dds_de)
    
    # print size factor
    size_factor = dds_de@colData@listData$sizeFactor    # the size factor for each replica
    print(size_factor)
    DESeq2_norm = counts(dds_de, normalized=TRUE)    # obtain the normalized counts for each replica
    normfile = paste(csvdir, "/", "norm_count.csv", sep='')
    write.csv(as.data.frame(DESeq2_norm), file=normfile)
    
    
    for (j in c(2:length(renames))){
      res = lfcShrink(dds_de, coef=j, type="apeglm")
      # create vocano plot
      myvocano(res, renames[j], vocadir, reflist[i], p_bar = 0.05)
      # create csv files
      resultcsv(res, renames[j], csvdir, reflist[i], p_bar = 0.05 , getall=1)
      summarylog(res, renames[j], csvdir, reflist[i], p_bar = 0.05 , getall=1)
    }
  }
  names(dds_de_list) = reflist
  return(dds_de_list)
}
```

```{r}
vofolder = "./DE-Volcano"
csvfolder = "./DE-CSVs"
ddsObj <- DESeqDataSetFromMatrix(countData = counts_data_qc,
                              colData = handle,
                              design = ~ condition_all)
de_object = deseqloop(ddsObj, reflist, vofolder, csvfolder)
```
```{r}
de_object
```

```{r}
i = 1
while (i<=length(de_object)) {
  
  j = 2
  while(j<=length(resultsNames(de_object[[i]]))){
    resLFC <- lfcShrink(de_object[[i]], coef=j, type="apeglm")
    plotMA(resLFC, alpha = 0.1, main = paste(reflist[i], resultsNames(de_object[[i]])[[j]], sep = "" ))
    j = j+1
  }
  
  i = i+1
}
```

```{r}
resLFC <- lfcShrink(de_object[[1]], coef=2, type="apeglm")
plotMA(resLFC)
```

